---
title: "How Important is it to Succeed vs. Good Teams in the Regular Season?"
author: "Rivka Boord"
date: '2022-09-08'
output:
  html_document:
    code_folding: none
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 2
  pdf_document:
    toc: yes
    toc_depth: '2'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = FALSE,
    eval = FALSE,
    message = FALSE,
    warning = FALSE,
    dpi = 300,
    fig.keep = 'all',
    fig.align = "center"
    )

# See Options: knitr::opts_chunk$get()
```

```{r}
library(baseballr)
library(rvest)
library(tidyverse)
```


```{r, eval=FALSE}
pull_record_vs_winning_teams <- function(year) {
    
    standings <- read_html(paste0('https://www.baseball-reference.com/leagues/majors/', year, '-standings.shtml')) %>% 
        html_nodes(xpath = '//comment()') %>% 
        html_text() %>%   
        paste(collapse = '') %>%   
        read_html() %>%   
        html_nodes('table') %>%    
        html_table() %>% 
        .[[2]] %>% 
        tibble()
    
    rec_vs_winning_teams <- standings %>% 
        rename(rec_vs_winning_teams = `=.500`) %>% 
        select(Tm, W, L, rec_vs_winning_teams) %>% 
        mutate(year = year)
    
    return(rec_vs_winning_teams)
    
}
```

```{r}

# scraping just the expanded standings for 2022 from Baseball Reference

expanded_baseball_standings_22 <- read_html('https://www.baseball-reference.com/leagues/majors/2022-standings.shtml') %>% 
    html_nodes(xpath = '//comment()') %>% 
    html_text() %>%
    paste(collapse = '') %>%
    read_html() %>%   
    html_nodes('table') %>% 
    html_table() %>% 
    .[[1]] %>% 
    tibble() %>% 
    rename(rec_vs_winning_teams = `=.500`)
```

## Find the record vs. winning teams for teams currently in 2022 playoff position

```{r}
current_playoff_teams_rec_vs_winning_teams <- expanded_baseball_standings_22 %>% 
    head(12) %>% 
    select(Tm, W, L, rec_vs_winning_teams) %>% 
    separate(
        col = rec_vs_winning_teams,
        into = c('wins_vs_winning_teams', 'losses_vs_winning_teams'),
        sep = '-',
        remove = TRUE
    ) %>% 
    mutate(wins_vs_winning_teams = as.numeric(wins_vs_winning_teams),
           losses_vs_winning_teams = as.numeric(losses_vs_winning_teams),
           total_gms_vs_winning_teams = (wins_vs_winning_teams + losses_vs_winning_teams),
           win_pct_vs_winning_teams = (wins_vs_winning_teams / total_gms_vs_winning_teams) %>% round(digits = 3)) %>% 
    arrange(desc(total_gms_vs_winning_teams))
```

## Find the list of all postseason teams since 1996, the first full season of Wild Card Round play

```{r, eval=FALSE}
library(Lahman)

Teams <- Teams %>%
    select(yearID, teamID, franchID, teamIDBR)

TeamsFranchises <- TeamsFranchises %>% 
    select(franchID, franchName)

team_id <- Teams %>% 
    left_join(TeamsFranchises, by = c('franchID'))

postseason_teams <- SeriesPost %>% 
    left_join(team_id, by = c('teamIDwinner' = 'teamID', 'yearID')) %>% 
    left_join(team_id, by = c('teamIDloser' = 'teamID', 'yearID')) %>% 
    rename(loser = franchName.y,
           winner = franchName.x) %>% 
    select(yearID, round, winner, lgIDwinner, loser, lgIDloser, wins, losses, ties) %>% 
    mutate(
        winner = case_when(
            yearID >= 2012 ~ str_replace_all(winner, 'Florida', 'Miami'),
        TRUE ~ winner
        ) %>% 
            str_replace_all('Metropolitans', 'Mets'), 
        loser = case_when(
            yearID >= 2012 ~ str_replace_all(loser, 'Florida', 'Miami'),
            TRUE ~ loser
               ) %>%
            str_replace_all('Metropolitans', 'Mets') 
           ) %>%
    rename(year = yearID) %>% 
    filter(year >= 1996)
```

## Standings from 1996-2021

```{r, eval=FALSE}
standings_1996_2021 <- map(c(1996:2021), pull_record_vs_winning_teams) %>% 
    data.table::rbindlist()

standings_1996_2021_cleaned <- standings_1996_2021 %>% 
    separate(
        col = rec_vs_winning_teams,
        into = c('wins_vs_winning_teams', 'losses_vs_winning_teams'),
        sep = '-',
        remove = TRUE
    ) %>% 
    mutate(
        wins_vs_winning_teams      = as.numeric(wins_vs_winning_teams),
        losses_vs_winning_teams    = as.numeric(losses_vs_winning_teams),
        total_gms_vs_winning_teams = (wins_vs_winning_teams + losses_vs_winning_teams),
        win_pct_vs_winning_teams   = (wins_vs_winning_teams / total_gms_vs_winning_teams) %>% round(digits = 3),
        Tm                         = Tm %>%  str_replace_all('Anaheim Angels', 'Los Angeles Angels of Anaheim'),
        year                       = as.numeric(year))
        ) %>%
    filter(Tm != 'Average')
```

## WS winners' record vs. winning teams

```{r}

ws_winners_record_vs_winning_teams <-  postseason_teams %>% 
    filter(round == 'WS') %>% 
    select(year, round, winner) %>% 
    left_join(standings_1996_2021_cleaned, by = c('winner' = 'Tm', 'year' = 'year')) %>% 
    summarize(
        mean_wins                          = mean(W) %>% round(1),
        mean_wins_vs_winning_teams         = mean(wins_vs_winning_teams) %>% round(1),
        mean_losses_vs_winning_teams       = mean(losses_vs_winning_teams) %>% round(1),
        mean_total_games_vs_winning_teams  = mean(total_gms_vs_winning_teams) %>% round(1),
        mean_win_pct_vs_winning_teams      = mean(win_pct_vs_winning_teams) %>% round(3),
        worst_win_pct_vs_winning_tms       = min(win_pct_vs_winning_teams),
        winning_recs_vs_winning_teams      = sum(win_pct_vs_winning_teams >= .500),
        losing_recs_vs_winning_teams       = sum(win_pct_vs_winning_teams < .500)
        ) %>%
    mutate(round = 'won WS')
```

## WS losers' record vs. winning teams

```{r}

lcs_winners_record_vs_winning_teams <-  postseason_teams %>% 
    filter(round == 'WS') %>% 
    select(year, round, loser) %>% 
    left_join(standings_1996_2021_cleaned, by = c('loser' = 'Tm', 'year' = 'year')) %>% 
    summarize(
        mean_wins                          = mean(W) %>% round(1),
        mean_wins_vs_winning_teams         = mean(wins_vs_winning_teams) %>% round(1),
        mean_losses_vs_winning_teams       = mean(losses_vs_winning_teams) %>% round(1),
        mean_total_games_vs_winning_teams  = mean(total_gms_vs_winning_teams) %>% round(1),
        mean_win_pct_vs_winning_teams      = mean(win_pct_vs_winning_teams) %>% round(3),
        worst_win_pct_vs_winning_tms       = min(win_pct_vs_winning_teams),
        winning_recs_vs_winning_teams      = sum(win_pct_vs_winning_teams >= .500),
        losing_recs_vs_winning_teams       = sum(win_pct_vs_winning_teams < .500)
        ) %>%
    mutate(round = 'won LCS')
```


## LCS losers' record vs. winning teams

```{r}
lds_winners_record_vs_winning_teams <- postseason_teams %>% 
    filter(round == 'ALCS' | round == 'NLCS') %>% 
    select(year, round, loser) %>% 
    left_join(standings_1996_2021_cleaned, by = c('loser' = 'Tm', 'year' = 'year')) %>% 
    mutate(winners_vs_winning_teams = sum(win_pct_vs_winning_teams >= .500)) %>% 
    summarize(
        mean_wins                          = mean(W) %>% round(1),
        mean_wins_vs_winning_teams         = mean(wins_vs_winning_teams) %>% round(1),
        mean_losses_vs_winning_teams       = mean(losses_vs_winning_teams) %>% round(1),
        mean_total_games_vs_winning_teams  = mean(total_gms_vs_winning_teams) %>% round(1),
        mean_win_pct_vs_winning_teams      = mean(win_pct_vs_winning_teams) %>% round(3),
        worst_win_pct_vs_winning_tms       = min(win_pct_vs_winning_teams),
        winning_recs_vs_winning_teams      = sum(win_pct_vs_winning_teams >= .500),
        losing_recs_vs_winning_teams       = sum(win_pct_vs_winning_teams < .500)
        ) %>%
    mutate(round = 'won LDS')
```

## LDS losers' record vs. winning teams

```{r}
lds_losers_record_vs_winning_teams <- postseason_teams %>% 
    filter(round %in% c('ALDS1', 'ALDS2', 'NLDS1', 'NLDS2')) %>% 
    select(year, round, loser) %>% 
        mutate(loser = case_when(year >= 2012 ~ str_replace_all(loser, 'Florida', 'Miami'),
                            TRUE ~ loser)) %>% 
    left_join(standings_1996_2021_cleaned, by = c('loser' = 'Tm', 'year' = 'year')) %>% 
    summarize(
        mean_wins                          = mean(W) %>% round(1),
        mean_wins_vs_winning_teams         = mean(wins_vs_winning_teams) %>% round(1),
        mean_losses_vs_winning_teams       = mean(losses_vs_winning_teams) %>% round(1),
        mean_total_games_vs_winning_teams  = mean(total_gms_vs_winning_teams) %>% round(1),
        mean_win_pct_vs_winning_teams      = mean(win_pct_vs_winning_teams) %>% round(3),
        worst_win_pct_vs_winning_tms       = min(win_pct_vs_winning_teams),
        winning_recs_vs_winning_teams      = sum(win_pct_vs_winning_teams >= .500),
        losing_recs_vs_winning_teams       = sum(win_pct_vs_winning_teams < .500)
        ) %>%
    mutate(round = 'lost LDS')
```

## find the differences in win rates vs. winning teams for each round of the playoffs

```{r}

postseason_teams_rec_vs_winning_teams_summarized <-  
    bind_rows(ws_winners_record_vs_winning_teams,
              lcs_winners_record_vs_winning_teams,
              lds_winners_record_vs_winning_teams,
              lds_losers_record_vs_winning_teams) %>% 
    mutate(
        pct_winners_vs_winning_teams = (winning_recs_vs_winning_teams /
            (winning_recs_vs_winning_teams + losing_recs_vs_winning_teams)) %>% round(3)
        ) %>% 
    select(round, everything())

postseason_teams_rec_vs_winning_teams_summarized
```
